Finiteness in Cubical Type Theory

### 2020-06-25 sections/search.tex
more on omniscience

So our task is to rewrite proof obligations like the one in
Equation~\ref{bool-assoc-auto-proof} into ones which can reduce completely. 
As it turns out, we have already described the type of proofs which can ``reduce
completely'': \emph{decidable} proofs.
If we have a decision procedure over some proposition \(P\) we can run that
decision during type checking, because the decision procedure itself is a proof
that the decision will terminate.
In code, we capture this idea with the following pair of functions:
\begin{multicols}{2}
  \begin{agdalisting}
    \ExecuteMetaData[agda/Snippets/Bool.tex]{is-true}
  \end{agdalisting} \columnbreak
  \begin{agdalisting}
    \ExecuteMetaData[agda/Snippets/Bool.tex]{from-true}
  \end{agdalisting}
\end{multicols}
The first is a function which derives a type from whether a decision is
successful or not.
This function is important because if we use the output of this type at any
point we will effectively force the unifier to run the decision computation.
The second takes---as an implicit argument---an inhabitant of the type generated
from the first, and uses it to prove that the decision can only be true, and the
extracts the resulting proof from that decision.
All in all, we can use it like this:
\begin{agdalisting}
  \ExecuteMetaData[agda/Snippets/Bool.tex]{extremely-obvious}
\end{agdalisting}
This technique will allow us to automatically compute any decidable predicate.
\section{Omniscience}
So we now know what is needed of us for proof automation: we need to take our
proofs and make them decidable.
In particular, our strategy will be to take a proposition which is decidable in
the individual case (like
\((x\;\AgdaFunction{∧}\;y)\;\AgdaFunction{∧}\;z\;\AgdaFunction{≡}\;x\;\AgdaFunction{∧}\;(y\;\AgdaFunction{∧}\;z)\)
for a given \(x\), \(y\), and \(z\)) and make it decidable in the universal case.
Whether this is possible or not is a property of the \emph{domain} of a
decidable predicate, and is described formally by the notion of ``Exhaustibility''.
\begin{definition}[Exhaustibility]
  We say a type \(A\) is exhaustible if, for any decidable predicate \(P\) on
  \(A\), the universal quantification of the predicate is decidable.
  \begin{equation}
    \left( \Pi {(x : A)} , \mathbf{Dec}(P(x)) \right) \rightarrow \mathbf{Dec} \left( \Pi {(x : A)} , P(x) \right)
  \end{equation}
\end{definition}
Perhaps surprisingly, it is not only finite types which can are exhaustible.
Certain infinite types can be
exhaustible~\cite{escardoInfiniteSetsThat2007}.

First we will explore the theoretical underpinning for search, and its relation
to finiteness.
The property we are looking for is effectively the ability to exhaustively
search through some type.
Consider the proof above in Equation~\ref{bool-assoc-auto-proof}: we know the
proof could be automatic because we can check every configuration of the values
of \(x\), \(y\), and \(z\).

\begin{definition}[Limited Principle of Omniscience]
  For any type \(A\) and predicate \(P\) on \(A\), the limited principle of
  omniscience \cite{myhillErrettBishopFoundations1972} is as follows:
  \begin{equation}
    \left( \Pi {(x : A)} , \mathbf{Dec}(P(x)) \right) \rightarrow \mathbf{Dec} \left( \Sigma {(x : A)} , P(x) \right)
  \end{equation}
  In other words, for any decidable predicate the existential quantification of
  that predicate is also decidable.
\end{definition}
The limited principle of omniscience is non-constructive, but individual types
can themselves satisfy omniscience.
In particular, cardinal finite types are omniscient.

There is also a universal form of omniscience, which we call exhaustibility.

All of the finiteness predicates we have seen imply exhaustibility.
Omniscience is stronger than exhaustibility, as we can derive the latter from
the former.
All of the ordered finiteness predicates imply omniscience.
For the unordered finiteness definitions, we have omniscience for prop-valued
predicates.

>>>>

So our task is to rewrite proof obligations like the one in
Equation~\ref{bool-assoc-auto-proof} into ones which can reduce completely. 
As it turns out, we have already described the type of proofs which can ``reduce
completely'': \emph{decidable} proofs.
If we have a decision procedure over some proposition \(P\) we can run that
decision during type checking, because the decision procedure itself is a proof
that the decision will terminate.
In code, we capture this idea with the following pair of functions:
\begin{multicols}{2}
  \begin{agdalisting}
    \ExecuteMetaData[agda/Snippets/Bool.tex]{is-true}
  \end{agdalisting} \columnbreak
  \begin{agdalisting}
    \ExecuteMetaData[agda/Snippets/Bool.tex]{from-true}
  \end{agdalisting}
\end{multicols}
The first is a function which derives a type from whether a decision is
successful or not.
This function is important because if we use the output of this type at any
point we will effectively force the unifier to run the decision computation.
The second takes---as an implicit argument---an inhabitant of the type generated
from the first, and uses it to prove that the decision can only be true, and the
extracts the resulting proof from that decision.
All in all, we can use it like this:
\begin{agdalisting}
  \ExecuteMetaData[agda/Snippets/Bool.tex]{extremely-obvious}
\end{agdalisting}
This technique will allow us to automatically compute any decidable predicate.
\section{Omniscience}
So we now know what is needed of us for proof automation: we need to take our
proofs and make them decidable.
In particular, our strategy will be to take a proposition which is decidable in
the individual case (like
\((x\;\AgdaFunction{∧}\;y)\;\AgdaFunction{∧}\;z\;\AgdaFunction{≡}\;x\;\AgdaFunction{∧}\;(y\;\AgdaFunction{∧}\;z)\)
for a given \(x\), \(y\), and \(z\)) and make it decidable in the universal case.
Whether this is possible or not is a property of the \emph{domain} of a
decidable predicate, and is described formally by the notion of ``Exhaustibility''.
\begin{definition}[Exhaustibility]
  We say a type \(A\) is exhaustible if, for any decidable predicate \(P\) on
  \(A\), the universal quantification of the predicate is decidable.
  \begin{equation}
    \left( \Pi {(x : A)} , \mathbf{Dec}(P(x)) \right) \rightarrow \mathbf{Dec} \left( \Pi {(x : A)} , P(x) \right)
  \end{equation}
\end{definition}
This property of \AgdaDatatype{Bool} would allow us to automate the proof of
associativity, but it is in fact not strong enough to find individual
representatives of a type which support some property.
For that we need the more well-known, but related, property of
\emph{omniscience}.
\begin{definition}[Limited Principle of Omniscience]
  For any type \(A\) and predicate \(P\) on \(A\), the limited principle of
  omniscience \cite{myhillErrettBishopFoundations1972} is as follows:
  \begin{equation}
    \left( \Pi {(x : A)} , \mathbf{Dec}(P(x)) \right) \rightarrow \mathbf{Dec} \left( \Sigma {(x : A)} , P(x) \right)
  \end{equation}
  In other words, for any decidable predicate the existential quantification of
  that predicate is also decidable.
\end{definition}

Both of these properties only make sense to talk about in a constructive
setting: classically speaking, \emph{every} type is omniscient and exhaustible.
Finite types are omniscient (the reason should be clear, but we will prove this
fact presently), but perhaps surprisingly, it is not \emph{only} finite types
which are exhaustible.
Certain infinite types can be
exhaustible~\cite{escardoInfiniteSetsThat2007}, but an exploration of that is
beyond the scope of this work.

Omniscience and exhaustibility are not interchangeable: conceptually, for
omniscience you need some kind of ordering on the underlying type.
If we had a property, say like the associativity property above, that is
satisfied by more than one element of the type, we would have to have some way
to pick a candidate element from the type to return in the case of the property
being satisfied.
This requirement of order mirrors the Ord difference in
Fig.~\ref{finite-classification}.

Omniscience implies exhaustibility, then, but the converse is not true.
\begin{lemma}
  Omniscience implies exhaustibility
\end{lemma}
\begin{proof}
  \todo{proof here}
\end{proof}

And the relation to the finiteness predicates is straightforward: all of the
finiteness predicates we have seen imply exhaustibility, and all of the ordered
finiteness predicates imply omniscience.
We can prove this by showing exhaustibility and omniscience for the weakest
candidate of finiteness predicates.
\begin{lemma}
  Kuratowski finiteness implies exhaustibility
\end{lemma}
\begin{lemma}
  Manifest enumerability implies omniscience
\end{lemma}

Finally, we can get around the order requirement for prop-valued predicates for
omniscience.
\begin{lemma}
  Omniscience and exhaustibility coincide for prop-valued predicates.
\end{lemma}
